On multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2022-07-29 paper/mcd2.tex
allow the obstacle to not have zero boundary values, so monotone restriction acts on general (non-zero boundary) FE functions

Let $\mathcal{V}^j$ be the continuous, piecewise-polynomial FE space $P_1$,\footnote{The FE method herein apparently extends to $Q_1$ spaces over quadrilateral meshes \cite{Elmanetal2014}.  However, for $k>1$ the $P_k$ space seems not to be suitable because the monotone restriction operator, \eqref{eq:fe:monotonerestriction} below, depends on element-wise monotonicity.  That is, for $P_1$ or $Q_1$ if the nodal values of an FE function (non-strictly) increase then the FE function also increases.  As shown by Figure 1.7 of \cite{Elmanetal2014}, for example, such monotonicity is not true for $P_2$.} over $\mathcal{T}^j$, with zero boundary values.  Note $\mathcal{V}^j$ is a subspace of $\mathcal{V}=W_0^{1,p}(\Omega)$ for any $p\ge 1$ \cite{Elmanetal2014}.  Our choices imply the following nesting:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}.  \label{eq:fe:nestedspaces}
\end{equation}
Suppose $\gamma^J \in \mathcal{V}^J$ denotes the \emph{obstacle} and define
\begin{equation}
\mathcal{K}^J = \left\{v \in \mathcal{V}^J\,:\,v \ge \gamma^J\right\}, \label{eq:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of \emph{admissible} FE functions.

One might assume that $f$ in \eqref{eq:vi} is defined over $\mathcal{K}^J$, i.e.~$f:\mathcal{K}^J \to \mathcal{V}'$.  However, especially for nontrivial nonlinear functionals, the finest-level, discretized FE functional $f^J:\mathcal{K}^J \to (\mathcal{V}^J)'$ is likely to only approximate $f$ because of quadrature, or other implementation details.  The functional $\ell$ in \eqref{eq:vi} is defined and continuous over $\mathcal{V}^J$, thus $\ell\in (\mathcal{V}^J)'$, and we assume for simplicity that it can be computed exactly, but we denote $\ell^J \in (\mathcal{V}^J)'$ to indicate that it must be represented in the FE sense.

@@

While \eqref{eq:fe:vi} approximates \eqref{eq:vi}, even if quadrature and other implementation details are ignored it may not technically be conforming for \eqref{eq:vi}.  That is, despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require $\mathcal{K}^J \subset \mathcal{K}$.  Only the fine-level FE obstacle $\gamma^J$ needs to be defined, while $\mathcal{K}$ remains abstract.  In fact, if a continuum obstacle exists at all then admissibility with respect to its FE interpolant $\gamma^J$ might not imply continuum admissiblity.  These caveats apply in geophysical situations where the obstacle is given by meshed elevation data, the continuum limit of which is undefined \cite{Bueler2016}.

The multilevel algorithm in the next section, designed to solve \eqref{eq:fe:vi}, requires transfer operators between meshes.  We use four such operators, three of which are the standard and linear maps used for nonlinear multigrid methods \cite{Trottenbergetal2001}.  Table \ref{tab:transfers} will help the reader keep track.

\begin{table}
\begin{tabular}{lll}
\emph{name}  & \emph{action}  & \emph{linear?} \\ \hline
canonical prolongation        & $P:\mathcal{V}^{j-1}\to\mathcal{V}^j$ & \,\checkmark \\
canonical (dual) restriction  & $R:(\mathcal{V}^j)'\to(\mathcal{V}^{j-1})'$ & \,\checkmark \\
(nodal) injection restriction & $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$ & \,\checkmark \\
monotone restriction          & $\mR:\mathcal{V}^j\to\mathcal{V}^{j-1}$ &  
\end{tabular}

\medskip
\caption{Transfer operators used in this paper.}
\label{tab:transfers}
\end{table}

@@

To define injection $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$, observe that for any given values at the (interior) nodes of $\mathcal{T}^j$ there is a unique function in $\mathcal{V}^j$ with those values.  Also observe that, by our refinement construction, the nodes of $\mathcal{T}^{j-1}$ are also nodes of $\mathcal{T}^j$.  For $z\in\mathcal{V}^j$ we define the injection $\iR z$ as the element of $\mathcal{V}^{j-1}$ which has the same nodal values as $z$ at the nodes of $\mathcal{T}^{j-1}$.  Note that injection clearly preserves ordering, that is, if $w \le z$ then $\iR w \le \iR z$.

Finally, monotone restriction $\mR:\mathcal{V}^j\to\mathcal{V}^{j-1}$ is a nonlinear map which maximizes nodal values over the supports of coarse-mesh hat functions.  To make the definition precise, first denote the \emph{interior nodes} of triangulation $\mathcal{T}^j$ by $x_q^j$ for $q=0,\dots,m_j-1$, and then the corresponding hat function \cite{Elmanetal2014} by $\psi_q^j$.  Monotone restriction of $z\in \mathcal{V}^j$ is
\begin{equation}
  \mR z = \sum_{p=1}^{m_{j-1}} \max \{z_q \,:\, \psi_p^{j-1}(x_q^j) > 0\}\, \psi_p^{j-1}. \label{eq:fe:monotonerestriction}
\end{equation}
Recalling that the piecewise-linear functions in $\mathcal{V}^j$ have zero boundary values, note that if $z\le 0$ then $0\ge \mR z \ge z$ on $\Omega$.

@@

To make this precise, suppose $w^J \in \cK^J$ is an admissible fine-level iterate which approximates the solution $u^J$ of VI \eqref{eq:fe:vi}.  From this iterate we define the \emph{defect constraint} \cite{GraeserKornhuber2009}
\begin{equation}
\chi^J = \gamma^J - w^J,  \label{eq:fe:defectconstraintfcn}
\end{equation}
a nonpositive function.  The motivation here is that a corrected iterate is admissible, $w^J + z^J \ge \gamma^J$, if and only if the perturbation $z^J$ exceeds the defect constraint, $z^J \ge \chi^J$.  Next, for $j=J,J-1,\dots,1$ we apply monotone restriction $\mR$ to define the \emph{level defect constraints}:
\begin{equation}
\chi^{j-1} = \mR \chi^j. \label{eq:fe:chilevels}
\end{equation}
As observed after equation \eqref{eq:fe:monotonerestriction}, $0\ge \chi^{j-1} \ge \chi^j$.  A $J=4$ example is shown in Figure \ref{fig:chiphilevels}.

We also denote the differences between the level defect constraints by
\begin{equation}
\phi^j = \chi^j - \chi^{j-1}. \label{eq:fe:philevels}
\end{equation}
(This uses the convenience definition $\chi^{-1}=0$, thus $\phi^0=\chi^0$ in particular.)  These are the gaps between the $\chi^j$ shown in Figure \ref{fig:chiphilevels}.  Note the telescoping sum $\sum_{j=0}^J \phi^j = \chi^J$ decomposition.

>>>>

Let $\tilde{\mathcal{V}}^j$ be the continuous, piecewise-polynomial FE space $P_1$,\footnote{The FE method herein apparently extends to $Q_1$ spaces over quadrilateral meshes \cite{Elmanetal2014}.  However, for $k>1$ the $P_k$ space seems not to be suitable because the monotone restriction operator, \eqref{eq:fe:monotonerestriction} below, depends on element-wise monotonicity.  That is, for $P_1$ or $Q_1$ if the nodal values of an FE function (non-strictly) increase then the FE function also increases.  As shown by Figure 1.7 of \cite{Elmanetal2014}, for example, such monotonicity is not true for $P_2$.} over $\mathcal{T}^j$, with arbitrary boundary values.  Let $\mathcal{V}^j \subset \tilde{\mathcal{V}}^j$ be the subspace with zero boundary values.  For any $p\ge 1$, $\tilde{\mathcal{V}}^j$ is a subspace of $W^{1,p}(\Omega)$ and $\mathcal{V}^j$ is a subspace of $\mathcal{V}=W_0^{1,p}(\Omega)$ \cite{Elmanetal2014}.  Our choices imply the following nesting:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}.  \label{eq:fe:nestedspaces}
\end{equation}

Our only need for the arbitrary-boundary-value spaces $\tilde{\mathcal{V}}^j$ is for the constraint.  Suppose $\gamma^J \in \tilde{\mathcal{V}}^j$ denotes the \emph{obstacle}, a function satisfying $\gamma^J \le 0$ at boundary nodes of $\mathcal{T}^J$, and define
\begin{equation}
\mathcal{K}^J = \left\{v \in \mathcal{V}^J\,:\,v \ge \gamma^J\right\}, \label{eq:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of \emph{admissible} FE functions.  Note that admissible functions from $\mathcal{K}^J$ have zero boundary values.

One might assume that the nonlinear functional $f$ in \eqref{eq:vi} is defined over $\mathcal{K}^J$, i.e.~$f:\mathcal{K}^J \to \mathcal{V}'$.  However, especially for nontrivial nonlinear functionals, the finest-level, discretized FE functional $f^J:\mathcal{K}^J \to (\mathcal{V}^J)'$ is likely to only approximate $f$ because of quadrature or other implementation details.  The functional $\ell$ in \eqref{eq:vi} is defined and continuous over $\mathcal{V}^J$, thus $\ell\in (\mathcal{V}^J)'$, and we assume for simplicity that it can be computed exactly, but we denote $\ell^J \in (\mathcal{V}^J)'$ to indicate that it must also be given an FE representation.

@@

While \eqref{eq:fe:vi} approximates \eqref{eq:vi}, it may not technically be conforming for \eqref{eq:vi}.  That is, even if quadrature and other implementation details are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require $\mathcal{K}^J \subset \mathcal{K}$.  Only the fine-level FE obstacle $\gamma^J$ needs to be defined, while $\mathcal{K}$ remains abstract.  In fact, if a continuum obstacle exists then admissibility with respect to its FE interpolant $\gamma^J$ might not imply continuum admissiblity.  These caveats apply in geophysical situations where the obstacle is given by meshed elevation data, the continuum limit of which is undefined \cite{Bueler2016}.

The multilevel algorithm in the next section, designed to solve \eqref{eq:fe:vi}, requires transfer operators between meshes.  We use four such operators, three of which are standard, linear maps used for multigrid methods \cite{Trottenbergetal2001}.  Table \ref{tab:transfers} will help the reader keep track.

\begin{table}
\begin{tabular}{lll}
\emph{name}  & \emph{action}  & \emph{linear?} \\ \hline
canonical prolongation        & $P:\mathcal{V}^{j-1}\to\mathcal{V}^j$ & \,\checkmark \\
canonical (dual) restriction  & $R:(\mathcal{V}^j)'\to(\mathcal{V}^{j-1})'$ & \,\checkmark \\
(nodal) injection restriction & $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$ & \,\checkmark \\
monotone restriction          & $\mR:\tilde{\mathcal{V}}^j\to\tilde{\mathcal{V}}^{j-1}$ &
\end{tabular}

\medskip
\caption{Transfer operators used in this paper.  FE vector space $\mathcal{V}^j$ has zero boundary values while $\tilde{\mathcal{V}}^j$ has arbitrary boundary values. $\mR$ is a nonlinear map.}
\label{tab:transfers}
\end{table}

@@

To define injection $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$, observe that for any given values at the (interior) nodes of $\mathcal{T}^j$ there is a unique function in $\mathcal{V}^j$ with those values.  Also observe that, by our refinement construction, the nodes of $\mathcal{T}^{j-1}$ are also nodes of $\mathcal{T}^j$.  For $z\in\mathcal{V}^j$ we define the injection $\iR z$ as the element of $\mathcal{V}^{j-1}$ which has the same nodal values as $z$ at the nodes of $\mathcal{T}^{j-1}$.  Note that injection preserves ordering, that is, if $w \le z$ then $\iR w \le \iR z$.

Finally, monotone restriction $\mR:\tilde{\mathcal{V}}^j\to\tilde{\mathcal{V}}^{j-1}$ is defined as the nonlinear map which maximizes nodal values over the supports of coarse-mesh hat functions.  To make the definition precise, first denote the \emph{nodes} of triangulation $\mathcal{T}^j$ by $x_q^j$ for $q=0,\dots,\tilde{m}_j-1$, and then the corresponding hat functions \cite{Elmanetal2014} by $\psi_q^j$.  (Boundary nodes are included here.)  Monotone restriction of $z\in\tilde{\mathcal{V}}^j$ is
\begin{equation}
  \mR z = \sum_{p=1}^{\tilde m_{j-1}} \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}\, \psi_p^{j-1}. \label{eq:fe:monotonerestriction}
\end{equation}
Note that if $z\in\tilde{\mathcal{V}}^j$ then $z \le \mR z$.

@@

To make this precise, suppose $w^J \in \cK^J$ is an admissible fine-level iterate which approximates the solution $u^J$ of VI \eqref{eq:fe:vi}.  From this iterate we define the \emph{defect constraint} \cite{GraeserKornhuber2009}
\begin{equation}
\chi^J = \gamma^J - w^J,  \label{eq:fe:defectconstraintfcn}
\end{equation}
a nonpositive function in $\tilde{\mathcal{V}}^J$.  The motivation here is that a corrected iterate is admissible, $w^J + z^J \ge \gamma^J$, if and only if the perturbation $z^J$ exceeds the defect constraint, $z^J \ge \chi^J$.  Next, for $j=J,J-1,\dots,1$ we apply monotone restriction $\mR$ to define the \emph{level defect constraints},
\begin{equation}
\chi^{j-1} = \mR \chi^j. \label{eq:fe:chilevels}
\end{equation}
so that $\chi^j \le \chi^{j-1}$.  A $J=4$ example is shown in Figure \ref{fig:chiphilevels}.

We also denote the differences between the level defect constraints by
\begin{equation}
\phi^j = \chi^j - \chi^{j-1}, \label{eq:fe:philevels}
\end{equation}
and these are nonpositive functions.  (Definition \eqref{eq:fe:philevels} uses $\chi^{-1}=0$.)  Functions $\phi^j$ are the gaps between the $\chi^j$ shown in Figure \ref{fig:chiphilevels}, with $\phi^0=\chi^0$ in particular.  Note the telescoping sum decomposition $\sum_{j=0}^J \phi^j = \chi^J$.
